---
title: "Compare SKT and GAM"
author: "Mike Wessel"
date: "2023-04-17"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, echo=FALSE}

# See library statements below for needed packages that need to be installed if not already
# e.g.  install.packages("rkt",dependencies=TRUE) # package to perform seasonal kendall tau;

# open libraries
library(tidyverse)
library(haven)
library(rkt)
library(here)
library(forecast)
library(pracma)
library(xts)
library(lmtest)
library(EnvStats)
library(imputeTS)
library(car)
library(foreign)
library(wqtrends)
library(lubridate)
library(kableExtra)


### CLEAR WORKSPACE - EACH RUN REQUIRES CLEARING CACHE TO RESET LISTS ETC..
rm(list = ls())

```


Initial data analysis conducted using R script SKT GAM Comp Run.r which saves out a file called df_total2. That file contains the results of SKT and the GAM based method applied to 15 sites including fixed stations, and estuarine and lake probabilistic designs. All sites contained at least 60 nonmissing observations which is  part of the standard PC inclusion criteria. 

The analysis includes two types of GAM based runs, one on the raw data (i.e., not date averaged so e.g. W1 would have 4 observations for each date), and one using a date average which preserves the period of time when there were 9 sampling periods which were then reduced to 8. 

The SKT uses sample period median values for its test based on the current 8 sample period assignments used for the Pinellas County Report Card.  

The results below suggest there a total of 8 discrepancies which we will examine one at a time in alpha numeric order. 

All the following plots and analysis use the "Raw" data to express the data in its most basic form

```{r summ, include=TRUE, echo=FALSE}

setwd(here::here("./data"))
results<-readRDS("df_total2.RDS")

sumit<-as_tibble(readRDS("df_total2.RDS"))%>%
    select(station,param,final_td,mt_td,mt_raw_td,final_est)%>%
    rename(skt=final_td,GAM_Avg=mt_td,GAM_Raw=mt_raw_td,skt_direction=final_est)%>%
    kbl(caption = paste0("Trend Result Comparison"))%>%
   kable_styling()
sumit
```

# First Example

09-02	TN: The SKT suggests NO TREND while both GAM based tests suggest trend. 

This site was not found to contain significant autocorrelation according to the SKT methodology. 

As seen in the plots below there is a data gap that may be throwing off the GAM method

```{r ex1,inclue=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
# get working directory
setwd(here::here("./data"))

mod<-readRDS("formt_raw.RDS")%>%
  filter(station == "09-02" & Param == "TN")%>%
  anlz_gam(., trans = 'log10')
ylab <- "TN"
show_prdseries(mod, ylab = ylab)+ggtitle("Raw Data Plot with GAM smooth fit")
show_perchg(mod, baseyr = 2003, testyr = 2020, ylab = "Chlorophyll-a (ug/L)")

#season
metseason <- anlz_metseason(mod, metfun = mean, doystr = 1, doyend = 364, nsim = 10000)
#Uses reml instead of ml
mixed<- anlz_mixmeta(metseason, yrstr = 2003, yrend = 2020)
show_metseason(mod, doystr = 1, doyend = 364, yrstr = 2003, yrend = 2020, ylab = "TN (mg/L)")
show_sumtrndseason(mod, doystr = 1, doyend = 364, justify = 'left', win = 5:15)

```

# Second Example

19-08 TN: The SKT suggests a statistically significant DECREASING TREND while both GAM based tests suggest NO TREND. 

This site was not found to contain significant autocorrelation according to the SKT methodology. 

GAM produced a warning that the Transformation introduced infinite values in continuous y-axis. 

Again,neither the Raw or Averaged GAM methods suggest significant results.

The regression plot with the box and whiskers below suggests a DECREASING slope.  

```{r ex2,inclue=FALSE, echo=FALSE, message=FALSE}
# get working directory
setwd(here::here("./data"))

mod<-readRDS("formt_raw.RDS")%>%
  filter(station == "19-08" & Param == "TN")%>%
  anlz_gam(., trans = 'log10')
ylab <- "TN"
show_prdseries(mod, ylab = ylab)+ggtitle("Raw Data Plot with GAM smooth fit")
show_perchg(mod, baseyr = 2003, testyr = 2020, ylab = "TN (mg/L)")

#season
metseason <- anlz_metseason(mod, metfun = mean, doystr = 1, doyend = 364, nsim = 10000)
#Uses reml instead of ml
mixed<- anlz_mixmeta(metseason, yrstr = 2003, yrend = 2020)
show_metseason(mod, doystr = 1, doyend = 364, yrstr = 2003, yrend = 2020, ylab = "TN (mg/L)")
show_sumtrndseason(mod, doystr = 1, doyend = 364, justify = 'left', win = 5:15)

```

# Third Example

W1 Chlorophyll a: The SKT suggests a statistically significant INCREASING TREND while both GAM based tests suggest NO TREND. 

This site was not found to contain significant autocorrelation according to the SKT methodology. 

GAM produced a warning that the Transformation introduced infinite values in continuous y-axis.

Again neither the raw or averaged GAM methods suggest significant results. 

The regression plot with the box and whiskers below suggest a INCREASING slope. 

```{r ex3,inclue=FALSE, echo=FALSE, message=FALSE}
# get working directory
setwd(here::here("./data"))

mod<-readRDS("formt_raw.RDS")%>%
  filter(station == "W1" & Param == "Chl_a")%>%
  anlz_gam(., trans = 'log10')
ylab <- "Chl_a"
show_prdseries(mod, ylab = ylab)+ggtitle("Raw Data Plot with GAM smooth fit")
show_perchg(mod, baseyr = 2003, testyr = 2020, ylab = "Chlorophyll-a (ug/L)")

#season
metseason <- anlz_metseason(mod, metfun = mean, doystr = 1, doyend = 364, nsim = 10000)
#Uses reml instead of ml
mixed<- anlz_mixmeta(metseason, yrstr = 2003, yrend = 2020)
show_metseason(mod, doystr = 1, doyend = 364, yrstr = 2003, yrend = 2020, ylab = "Chl_a (ug/L)")
show_sumtrndseason(mod, doystr = 1, doyend = 364, justify = 'left', win = 5:15)

```


# Fourth Example

W4 TN: The results were Trend, Trend, No Trend for the SKT, Gam Avg, and Gam Raw, respectively.  

This site was not found to contain significant autocorrelation according to the SKT methodology. 


The percent change plot suggests a significant difference between start and end points.

The regression plot with the box and whiskers below suggest a increasing slope. 

The GAM averaged (and SKT) result was significant but the GAM Raw was not

```{r ex4,inclue=FALSE, echo=FALSE, message=FALSE}
# get working directory
setwd(here::here("./data"))

mod<-readRDS("formt_raw.RDS")%>%
  filter(station == "W4" & Param == "TN")%>%
  anlz_gam(., trans = 'log10')
ylab <- "TN"
show_prdseries(mod, ylab = ylab)+ggtitle("Raw Data Plot with GAM smooth fit")
show_perchg(mod, baseyr = 2003, testyr = 2020, ylab = "TN (mg/L)")

#season
metseason <- anlz_metseason(mod, metfun = mean, doystr = 1, doyend = 364, nsim = 10000)
#Uses reml instead of ml
mixed<- anlz_mixmeta(metseason, yrstr = 2003, yrend = 2020)
show_metseason(mod, doystr = 1, doyend = 364, yrstr = 2003, yrend = 2020, ylab = "TN (mg/L)")
show_sumtrndseason(mod, doystr = 1, doyend = 364, justify = 'left', win = 5:15)

```

# Fifth Example

W5 Chl_a: In this example the exact opposite occurs: No Trend, No Trend, Trend. 

The raw GAM fit suggest no seasonality - looks like a bad fit overall 

The percent change plot suggests NO significant difference between start and end points. 

However, the regression plot with the box and whiskers below suggest a increasing slope. 

```{r ex5,inclue=FALSE, echo=FALSE, message=FALSE}
# get working directory
setwd(here::here("./data"))

mod<-readRDS("formt_raw.RDS")%>%
  filter(station == "W5" & Param == "Chl_a")%>%
  anlz_gam(., trans = 'log10')
ylab <- "Chl_a"
show_prdseries(mod, ylab = ylab)+ggtitle("Raw Data Plot with GAM smooth fit")
show_perchg(mod, baseyr = 2003, testyr = 2020, ylab = "Chlorophyll-a (ug/L)")

#season
metseason <- anlz_metseason(mod, metfun = mean, doystr = 1, doyend = 364, nsim = 10000)
#Uses reml instead of ml
mixed<- anlz_mixmeta(metseason, yrstr = 2003, yrend = 2020)
show_metseason(mod, doystr = 1, doyend = 364, yrstr = 2003, yrend = 2020, ylab = "Chl_a (ug/L)")
show_sumtrndseason(mod, doystr = 1, doyend = 364, justify = 'left', win = 5:15)

```


# Sixth Example

W6 TN: The Results of this example suggest: Trend, Trend,  No Trend for SKT, GAM avg, and GAM raw, respectively

The percent change plot suggests a significant difference between start and end points.

The regression plot with the box and whiskers below suggest a increasing slope. 


```{r ex6,inclue=FALSE, echo=FALSE, message=FALSE}
# get working directory
setwd(here::here("./data"))

mod<-readRDS("formt_raw.RDS")%>%
  filter(station == "W6" & Param == "TN")%>%
  anlz_gam(., trans = 'log10')
ylab <- "TN"
show_prdseries(mod, ylab = ylab)+ggtitle("Raw Data Plot with GAM smooth fit")
show_perchg(mod, baseyr = 2003, testyr = 2020, ylab = "TN (mg/L)")

#season
metseason <- anlz_metseason(mod, metfun = mean, doystr = 1, doyend = 364, nsim = 10000)
#Uses reml instead of ml
mixed<- anlz_mixmeta(metseason, yrstr = 2003, yrend = 2020)
show_metseason(mod, doystr = 1, doyend = 364, yrstr = 2003, yrend = 2020, ylab = "TN (mg/L)")
show_sumtrndseason(mod, doystr = 1, doyend = 364, justify = 'left', win = 5:15)

```

# Seventh Example

W7 Chla Result: Trend,  No Trend, No Trend

The percent change plot suggests a significant difference between start and end points.

The regression plot with the box and whiskers below suggest a slight decreasing slope.


```{r ex7,inclue=FALSE, echo=FALSE, message=FALSE}
# get working directory
setwd(here::here("./data"))

mod<-readRDS("formt_raw.RDS")%>%
  filter(station == "W7" & Param == "Chl_a")%>%
  anlz_gam(., trans = 'log10')
ylab <- "Chl_a"
show_prdseries(mod, ylab = ylab)+ggtitle("Raw Data Plot with GAM smooth fit")
show_perchg(mod, baseyr = 2003, testyr = 2020, ylab = "Chlorophyll a (ug/L)")

#season
metseason <- anlz_metseason(mod, metfun = mean, doystr = 1, doyend = 364, nsim = 10000)
#Uses reml instead of ml
mixed<- anlz_mixmeta(metseason, yrstr = 2003, yrend = 2020)
show_metseason(mod, doystr = 1, doyend = 364, yrstr = 2003, yrend = 2020, ylab = "Chl_a (ug/L)")
show_sumtrndseason(mod, doystr = 1, doyend = 364, justify = 'left', win = 5:15)

```

# Eighth Example

W8 Chla Result: No Trend,  No Trend, Trend

Looks like a poor GAM fit overall, similar to example 5

The percent change plot suggests NO significant difference between start and end point.

The regression plot with the box and whiskers below suggest a increasing slope. 

```{r ex8,inclue=FALSE, echo=FALSE, message=FALSE}
# get working directory
setwd(here::here("./data"))

mod<-readRDS("formt_raw.RDS")%>%
  filter(station == "W8" & Param == "Chl_a")%>%
  anlz_gam(., trans = 'log10')
ylab <- "Chl_a"
show_prdseries(mod, ylab = ylab)+ggtitle("Raw Data Plot with GAM smooth fit")
show_perchg(mod, baseyr = 2003, testyr = 2020, ylab = "Chlorophyll a (ug/L)")

#season
metseason <- anlz_metseason(mod, metfun = mean, doystr = 1, doyend = 364, nsim = 10000)
#Uses reml instead of ml
mixed<- anlz_mixmeta(metseason, yrstr = 2003, yrend = 2020)
show_metseason(mod, doystr = 1, doyend = 364, yrstr = 2003, yrend = 2020, ylab = "Chl_a (ug/L)")
show_sumtrndseason(mod, doystr = 1, doyend = 364, justify = 'left', win = 5:15)

```


